---
title: "Two-way ANOVA with R"
author: "Stéphane Laurent"
date: "2017-10-19"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Data 

The data are simulated for this example.

```{r}
library(AOV2F)
set.seed(666)
dat <- SimDataAV2(I=3, J=4, Kmin=1, Kmax=3, p=c(0.1,0.2), 
                 factor.names=c("factor1", "factor2"), resp.name="y")
head(dat)
```

# Inputs 

```{r}
factors = c("factor1", "factor2") 
resp = "y"
factor.names = c("FACTOR1", "FACTOR2")
resp.name = "RESPONSE"
factor1.ref = "A3"
factor2.ref = "B1"
```

# Analysis

```{r echo=TRUE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(ztable)
library(ggplot2)
library(gridExtra)
library(lsmeans)
library(multcomp)
library(stringr)
library(plyr)
library(car)
options(knitr.table.format = "html") 
```


- Rename the column names and set the reference levels:

```{r echo=TRUE}
dat <- dat[, c(factors, resp)]
names(dat) <- c("factor1", "factor2", "resp")
factor1.name <- factor.names[1]
factor2.name <- factor.names[2]
dat$factor1 <- relevel(dat$factor1, ref=factor1.ref)
dat$factor2 <- relevel(dat$factor2, ref=factor2.ref)
```

## Data summaries 

- Check whether there are missing values:

```{r echo=TRUE, results='asis', comment=NA}
NAs <- length(which(is.na(dat$resp)))
dat <- droplevels(subset(dat, !is.na(resp)))
if(NAs > 0){
  if(NAs==1){
    cat("There was one missing value in the response column.")
  }else{
    cat(sprintf("There were %s missing values in the response column.", NAs))
  }
}else{
  cat("There is no missing value in the response column.")
}
```

- Plot all data:

```{r echo=TRUE, fig.height=nlevels(dat$factor1), fig.width=2*nlevels(dat$factor2)}
ggplot(data=dat, aes(x=resp, y=factor1)) +
  geom_point() + 
  geom_jitter(width=0, height=.2) + 
  ylab(factor1.name) + xlab(resp.name) +
  facet_grid(.~factor2) +
  ggtitle("Plot of data by each combination of the two factors")
```

- Table of means:

```{r echo=TRUE, results='asis'}
xfrequencies <- xtabs(~factor1+factor2, data=dat)
xmeans <- xtabs(resp~factor1+factor2, data=dat) /  xfrequencies
names(dimnames(xmeans)) <- factor.names
kable(format_ftable(ftable(xmeans)), 
             caption="Means by each combination of the two factors") %>% 
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
```

- Table of counts: 

```{r echo=TRUE}
names(dimnames(xfrequencies)) <- factor.names
kable(format_ftable(ftable(xfrequencies)), 
      caption="Frequencies by each combination of the two factors") %>% 
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
```

The two tables above are generated with the help of this function `format_ftable`:

```{r echo=TRUE, eval=FALSE}
format_ftable <- function (x, digits = getOption("digits"))
{
  if (!inherits(x, "ftable"))
    stop("x must be an `ftable'")
  ox <- x
  makeLabels <- function(lst) {
    lens <- sapply(lst, length)
    cplensU <- c(1, cumprod(lens))
    cplensD <- rev(c(1, cumprod(rev(lens))))
    y <- NULL
    for (i in rev(seq(along = lst))) {
      ind <- 1 + seq(from = 0, to = lens[i] - 1) * cplensD[i+1]
      tmp <- character(length = cplensD[i])
      tmp[ind] <- lst[[i]]
      y <- cbind(rep(tmp, times = cplensU[i]), y)
    }
    y
  }
  makeNames <- function(x) {
    nmx <- names(x)
    if (is.null(nmx))
      nmx <- rep("", length = length(x))
    nmx
  }
  xrv <- attr(x, "row.vars")
  xcv <- attr(x, "col.vars")
  LABS <- cbind(rbind(matrix("", nrow = length(xcv), ncol = length(xrv)),
                      makeNames(xrv), 
                      makeLabels(xrv)), 
                c(makeNames(xcv), rep("", times = nrow(x) + 1)))
  DATA <- rbind(t(makeLabels(xcv)), rep("", times = ncol(x)),
                format(unclass(x), digits = digits))
  cbind(apply(LABS, 2, format, justify = "left"), 
        apply(DATA, 2, format, justify = "right"))
}
```


- Plot of means for each combination of the two factors: 

```{r echo=TRUE, fig.height=1.33*nlevels(dat$factor1)}
means <- aggregate(resp~factor1+factor2, data=dat, FUN=mean)
ggplot(data=means, aes(y=resp, x=factor1, colour=factor2, group=factor2)) +
  geom_line(size=.6) + geom_point(size=3) +
  labs(colour=factor2.name) + 
  xlab(factor1.name) + ylab(resp.name) + coord_flip() +
  ggtitle("Means by each combination of the two factors")
```

- Plot of overall means:

```{r echo=TRUE, fig.width=12, fig.height=1.5*max(nlevels(dat$factor1),nlevels(dat$factor2))}
gg1 <- ggplot(aggregate(resp~factor1, data=dat, FUN=mean), 
              aes(y=resp, x=factor1, group=NA)) +
  geom_line(size=.6, colour="black") +
  geom_point(size=3, colour="forestgreen") +
  xlab(factor1.name) + ylab(resp.name) + coord_flip() +
  ggtitle(paste0("Overall means by factor ", factor1.name))
gg2 <- ggplot(aggregate(resp~factor2, data=dat, FUN=mean), 
              aes(y=resp, x=factor2, group=NA)) +
  geom_line(size=.6, colour="black") +
  geom_point(size=3, colour="forestgreen") +
  xlab(factor2.name) + ylab(resp.name) + coord_flip() +
  ggtitle(paste0("Overall means by factor ", factor2.name))
grid.arrange(gg1, gg2, ncol=2)
```


## Two-way ANOVA

- Fit the model:

```{r echo=TRUE, results='asis'}
lmfit <- lm(resp ~ factor1*factor2, data=dat)
interact <- anova(lmfit)$`Pr(>F)`[3] < 5/100
aovTable <- Anova(lmfit)
ztable(aovTable, type="html")
```

- Individual confidence intervals of the means:

```{r echo=TRUE, fig.height=1.33*nlevels(dat$factor1), fig.width=2*nlevels(dat$factor2)}
intervals <- confint(emmeans(lmfit, ~factor1+factor2), level=95/100)
ggplot(data=intervals, aes(x=factor1, y=emmean)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=.1) +
  xlab(factor1.name) + ylab(resp.name) +
  facet_grid(.~factor2) + coord_flip() +
  ggtitle("Individual confidence intervals of the means")
```


```{r echo=TRUE, results='asis'}
tab <- as.data.frame(intervals)
kable(tab, caption="Individual confidence intervals of the means") %>% 
    kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
```

- Simultaneous confidence intervals of the means:

```{r echo=TRUE, fig.height=1.33*nlevels(dat$factor1), fig.width=2*nlevels(dat$factor2), error=TRUE}
emm <- emmeans(lmfit, ~factor1+factor2)
K <- emm@linfct
tests <- glht(lmfit, linfct=K)
Intervals <- intervals <- 
  cbind(emm@grid, data.frame(confint(tests, level=0.95)$confint))
# if(factor1.ref %in% levels(ggd$factor1)){ # ???
#   ggd$factor1 <- relevel(ggd$factor1, ref=factor1.ref)
# }
ggplot(data=intervals, aes(x=factor1, y=Estimate)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=lwr, ymax=upr), width=.1) +
  xlab(factor1.name) + ylab(resp.name) +
  facet_grid(.~factor2) + coord_flip() +
  ggtitle("Simultaneous confidence intervals of the means")
```

```{r echo=TRUE, results='asis'}
kable(intervals, caption="Simultaneous confidence intervals of the means") %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
```


### Comparisons to the reference level of factor1 by level of factor 2

```{r echo=TRUE, fig.height=nlevels(dat$factor1), fig.width=2*nlevels(dat$factor2)}
if(factor1.ref %in% levels(Intervals$factor1)){
  C <- emmeans(lmfit, pairwise ~ factor1 | factor2, weights="cells")$contrasts
  M <- C@linfct
  indices <- !is.na(str_locate(C@grid$contrast, paste0(factor1.ref, " -"))[,"start"]==1)
  comparisons <- rename(droplevels(C@grid[indices,]), c(contrast="Comparison", factor2=factor2.name))
  M <- M[indices,]
  ci <- confint(glht(lmfit, linfct = M))
  intervals <- cbind(comparisons, data.frame(ci$confint))
  intervals$Comparison <- str_replace(intervals$Comparison, "-", "vs")
  intervals$signif <- intervals$lwr>0 | intervals$upr<0
  colors <- if(all(intervals$signif)) "red" else if(all(!intervals$signif)) "black" else c("black", "red")
  ggplot(intervals, aes(y=Estimate, x=Comparison, color=signif)) + geom_point(size=3) +
    geom_errorbar(aes(ymin=lwr, ymax=upr), width=.1) +
    facet_grid(as.formula(sprintf(".~%s", factor2.name))) + coord_flip() +
    geom_hline(yintercept=0, linetype="dashed")  +
    scale_colour_manual(values=colors) + guides(color=FALSE) +
    ylab("Mean difference")
}else{
  cat("These comparisons are not available because the reference level of factor1 is not here.")
}
```

```{r echo=TRUE, results='asis'}
if(factor1.ref %in% levels(Intervals$factor1)){
  tab <- intervals[,c("Comparison", factor2.name, "Estimate", "lwr", "upr")]
  kable(tab, 
        caption="Comparisons to the reference level of factor1 for each level of factor2") %>% 
      kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
}
```


### Comparisons to the reference level of factor 1 over all levels of factor 2 

```{r echo=TRUE, results='asis'}
if(interact && factor1.ref %in% levels(Intervals$factor1)) cat("NOTE: Results may be misleading due to the significant interaction !")
```

```{r echo=TRUE, message=FALSE, fig.height=nlevels(dat$factor1)}
if(factor1.ref %in% levels(Intervals$factor1)){
  ci <- emmeans(lmfit, pairwise ~ factor1, weights="cells")$contrasts
  M <- ci@linfct
  rownames(M) <- ci@grid$contrast
  if(nlevels(dat$factor1)>2)  M <- M[which(!is.na(str_locate(rownames(M), paste0(factor1.ref, " -"))[,"start"]==1)), ]
  intervals <- data.frame(confint(glht(lmfit, linfct = M))$confint)
  tab <- intervals <- cbind(Comparison=str_replace(rownames(intervals), "-", "vs"), intervals)
  intervals$signif <- intervals$lwr>0 | intervals$upr<0
  colors <- if(all(intervals$signif)) "red" else if(all(!intervals$signif)) "black" else c("black", "red")
  ggplot(intervals, aes(y=Estimate, x=Comparison, color=signif)) + 
    geom_point(size=3) +
    geom_errorbar(aes(ymin=lwr, ymax=upr), width=.1) +
    coord_flip() +
    geom_hline(yintercept=0, linetype="dashed")  +
    scale_colour_manual(values=colors) + guides(color=FALSE) +
    ylab("Mean difference")
}else{
  cat("These comparisons are not available because the reference level of factor1 is not here.")
}
```

```{r echo=TRUE, results='asis'}
if(factor1.ref %in% levels(Intervals$factor1)){
    kable(tab, caption="Comparisons to the reference level of factor1 over all levels of factor2", digits=2) %>%
    kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                  full_width = FALSE)
}
```


### Comparisons to the reference level of factor2 by level of factor 1

```{r echo=TRUE, fig.height=nlevels(dat$factor2), fig.width=2*nlevels(dat$factor1)}
if(factor2.ref %in% levels(Intervals$factor2)){
  C <- emmeans(lmfit, pairwise ~ factor2 | factor1, weights="cells")$contrasts
  M <- C@linfct
  indices <- !is.na(str_locate(C@grid$contrast, 
                               paste0(factor2.ref, " -"))[,"start"]==1)
  comparisons <- rename(droplevels(C@grid[indices,]), 
                        c(contrast="Comparison", factor1=factor1.name))
  M <- M[indices,]
  ci <- confint(glht(lmfit, linfct = M))
  intervals <- cbind(comparisons, data.frame(ci$confint))
  intervals$Comparison <- str_replace(intervals$Comparison, "-", "vs")
  intervals$signif <- intervals$lwr>0 | intervals$upr<0
  colors <- if(all(intervals$signif)) "red" else if(all(!intervals$signif)) "black" else c("black", "red")
  ggplot(intervals, aes(y=Estimate, x=Comparison, color=signif)) + 
    geom_point(size=3) +
    geom_errorbar(aes(ymin=lwr, ymax=upr), width=.1) +
    facet_grid(as.formula(sprintf(".~%s", factor1.name))) + coord_flip() +
    geom_hline(yintercept=0, linetype="dashed")  +
    scale_colour_manual(values=colors) + guides(color=FALSE) +
    ylab("Mean difference")
}else{
  cat("These comparisons are not available because the reference level of factor2 is not here.")
}
```

```{r echo=TRUE, results='asis'}
if(factor2.ref %in% levels(Intervals$factor2)){
  tab <- intervals[,c("Comparison", factor1.name, "Estimate", "lwr", "upr")]
  kable(tab, caption="Comparisons to the reference level of factor2 for each level of factor1") %>%
          kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                full_width = FALSE)
}
```



### Comparisons to the reference level of factor 2 over all levels of factor 1 

```{r echo=TRUE, results='asis'}
if(interact && factor2.ref %in% levels(Intervals$factor2)) cat("NOTE: Results may be misleading due to the significant interaction !")
```


```{r echo=TRUE, message=FALSE, fig.height=nlevels(dat$factor2)}
if(factor2.ref %in% levels(Intervals$factor2)){
  ci <- emmeans(lmfit, pairwise ~ factor2, weights="cells")$contrasts
  M <- ci@linfct
  rownames(M) <- ci@grid$contrast
  if(nlevels(dat$factor2)>2)  M <- 
    M[which(!is.na(str_locate(rownames(M), 
                              paste0(factor2.ref, " -"))[,"start"]==1)), ]
  intervals <- data.frame(confint(glht(lmfit, linfct = M))$confint)
  tab <- intervals <- cbind(Comparison=str_replace(rownames(intervals), "-", "vs"), intervals)
  intervals$signif <- intervals$lwr>0 | intervals$upr<0
  colors <- if(all(intervals$signif)) "red" else if(all(!intervals$signif)) "black" else c("black", "red")
  ggplot(intervals, aes(y=Estimate, x=Comparison, color=signif)) + geom_point(size=3) +
    geom_errorbar(aes(ymin=lwr, ymax=upr), width=.1) +
    coord_flip() +
    geom_hline(yintercept=0, linetype="dashed")  +
    scale_colour_manual(values=colors) + guides(color=FALSE) +
    ylab("Mean difference")
}else{
  cat("These comparisons are not available because the reference level of factor2 is not here.")
}
```

```{r echo=TRUE, results='asis'}
if(factor2.ref %in% levels(Intervals$factor2)){
  kable(tab, caption="Comparisons to the reference level of factor2 over all levels of factor1") %>%
    kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), 
                  full_width = FALSE)
}
```

